Optifcal Observations of an Extraordinarily Young Type Ia Supernova iPTF16abc

### 2017-05-12 doc/ms.tex
backup

Although Type Ia supernovae (SNe Ia) have been extensively used as
standardizable candles, their progenitor scenarios and explosion
physics are still in debate (see a recent review by
\citealt{2014ARA&A..52..107M}). Detailed extermely early-phase
observations are one of the most promising avenues to further
constrain this problem.

While the shock breakout of a SN Ia occurs on a sub-second timescale,
the subsequent quasi-adiabatic expanding and cooling of the unbinded
ejecta produces thermal emissions that can be used to infer the
original size of the exploding star
\citep{2010ApJ...708..598P,2011ApJ...728...63R}. Comparing models of
these cooling emissions to the earliest-phase data of SN2011fe,
\citet{2012ApJ...744L..17B} concluded that the radius of the
progenitor star is $\lesssim0.01\sr$ where $\sr$ is the solar
radius. Combining this size constraint and the measured ejecta mass to
derive the mean density of the progenitor star, we confirmed that the
progenitor star is compact and degenerate. Admittedly, due to the
initial small surface area of the progenitor star, the shock cooling
emission of a SN Ia decays drastically as the ejecta expands. Given
typical parameters of a SN Ia, this thermal emission is visible from
events up to $\sim 10\,\textrm{Mpc}$ within one day of their explosions.

Another expectation from the extremely early-phase observations of a
SN Ia is the excess emission from collisions between SN ejecta and a
companion star, a natural consequence from the single-degenerate
progenitor hypothesis \citep{1973ApJ...186.1007W,2010ApJ...708.1025K}.
In a low-velocity SN Ia iPTF14atg, \citet{2015Natur.521..328C} for the
first time detected a strong and declining ultraviolet pulse within a
few days of the SN explosion which is best interpreted as the
SN-companion collision. This signature has been searched in a number
of nearby, early-phase normal SNe Ia, but most of these studies result
in no detection
\citep{2010ApJ...722.1691H,2011ApJ...741...20B,2012ApJ...744...38F,
  2012ApJ...744L..17B,2015Natur.521..332O,
  2013ApJ...778L..15Z,2015ApJ...799..106G,2016ApJ...826..144S,
  2015ApJS..221...22I}. The exception is SN2012cg which
\citet{2016ApJ...820...92M} claimed detection of blue excess in the
earliest-phase light curve and attribute it to SN-companion collision.
However, this statement is recently challenged by
\citet{2016arXiv161007601S}. In fact, it is not surprising that no
SN-companion collision has been observationally confirmed, because
only up to $\sim 10\%$ of events from the single-degenerate channel
have the preferred binary geometry for us to see the collision
signatures.

More commonly, the only observed light curve of a SN Ia is purely
powered by the radioactive decay of synthesized $^{56}$Ni. Since
$^{56}$Ni atoms are synthesized, mixed and deposited into different
layers inside the ejecta, the SN may experience a dark period after
the SN shock breakout and before the radioactive energy diffuses to
the photosphere \citep{2014ApJ...784...85P}. In the case of strong
mixing, $^{56}$Ni energy can reach the photosphere rapidly. As such,
the dark period is very short and the light curve shows a fast initial
rise. In the opposite case, it may take up to a couple of days for
the $^{56}$Ni energy to diffuse to the photosphere. The initial rise
of the light curve is then moderate \citep{2016ApJ...826...96P}. To
summarize, the initial rise of the light curve of a SN Ia conveys
information on distribution of synthesized $^{56}$Ni.

In this paper, we report observations of an extraordinarily young SN
Ia iPTF16abc, which was discovered by the intermediate Palomar
Transient Factory on 2016 April $3.36$\footnote{all times in this
  paper are in UTC.} at $\textrm{R.A.}=13^h34^m45.49^s$,
$\textrm{Dec.}=+13^d51^m14.3^s$ (J2000) with a $g$-band magnitude of
$21.31\pm0.27$ \citep{2016PASP..128k4502C,2016ATel.8907....1M}. The
transient is spatially coincident with a tidal tail of the galaxy
NGC\,5221 at 100\,Mpc. No activity was detected at the same location
down to $g=22.1$\,mag on April $2.42$. Our spectroscopic follow-up
compaign classified iPTF16abc as a normal SN Ia
\citep{2016ATel.8909....1C}.

>>>>

Although Type Ia supernovae (SNe Ia) have been extensively used as
standardizable candles, their progenitor systems and explosion
mechanisms remain unclear (see a recent review by
\citealt{2014ARA&A..52..107M}). Observations of extremely young
events are one promising avenue to further
investigation of this problem.

While the shock breakout of a SN Ia occurs on a minisecond timescale,
the subsequent adiabatic cooling of the unbinded ejecta during its free
expansion produces thermal emission, before the radioactive energy manages
to heat up the photosphere of the ejecta. The thermal emission during this
phase therefore be used to infer the original size of the exploding star
\citep{2010ApJ...708..598P,2011ApJ...728...63R}. For example,
the earliest observations of SN2011fe allowed
\citet{2012ApJ...744L..17B} to conclude that the radius of the
progenitor star is $\lesssim0.01\sr$ where $\sr$ is the solar
radius and that the progenitor has to be compact and degenerate.
However, due to the small surface area of the progenitor star,
the cooling emission at this phase is not luminous and decays rapidly.
As a result, this method is only applicable to events
within $\sim 10\,\textrm{Mpc}$.

If a SN Ia is born in the single degenerate (SD) channel
\citep{1973ApJ...186.1007W}, then the SN ejecta slams into the companion and
produces a non-thermal X-ray flash. Once the fast-moving ejecta covers the
collision region, the collision heats up the surrounding material. The
collisional energy is then released near the direction of the companion (about
a few to ten percent of the $4\pi$ solid angle) as thermal emission with a
spectrum that peaks in the UV \citep{2010ApJ...708.1025K}.
This SN-companion collisional
signature serves as a ``smoking gun'' for the SD channel, because it is not
expected in the competing double-degenerate channel \ycao{REF}. This signature
has been searched in several nearby SNe Ia and a few samples of distant
SNe Ia \citep{2010ApJ...722.1691H,2011ApJ...741...20B,2012ApJ...744...38F,
2012ApJ...744L..17B,2015Natur.521..332O,2013ApJ...778L..15Z,
2015ApJ...799..106G,2016ApJ...826..144S,2015ApJS..221...22I}, only in a
low-velocity SNe Ia was a strong and declining ultraviolet (UV) pulse seen
within a few days of the explosion.

More commonly, the observed light curve of a SN Ia is purely
powered by the radioactive decay of synthesized $^{56}$Ni.
Depending on the depth of the shallowest layer where $^{56}$Ni is deposited,
the SN may experience a dark period before the radioactive energy diffuses to
the photosphere. According to the calculation in \citet{2016ApJ...826...96P},
strong mixing leads to a short dark period and a fast initial rise, while
weak mixing results in a long dark period and a moderate initial rise.
Measuring the initial rise rate of a SN Ia allows us to constrain the mixing
level of the SN ejecta which is closely related to the explosion mechanism.

In this paper, we report observations of a young SN Ia iPTF16abc within a day
of its explosion. This SN was discovered by the intermediate Palomar
Transient Factory on 2016 April $3.36$\footnote{all times in this
  paper are in UTC.} at $\textrm{R.A.}=13^h34^m45.49^s$,
$\textrm{Dec.}=+13^d51^m14.3^s$ (J2000) with a $g$-band magnitude of
$21.31\pm0.27$ \citep{2016PASP..128k4502C,2016ATel.8907....1M}. The
transient is spatially coincident with a tidal tail of the galaxy
NGC\,5221 at 100\,Mpc. No activity was detected at the same location
down to $g=22.1$\,mag on April $2.42$. Our spectroscopic follow-up
compaign classified iPTF16abc as a normal SN Ia \citep{2016ATel.8909....1C}.
